Equalizer Programming Guide

### 2008-01-15 Developer/ProgrammingGuide/paper.tex
Update Programmers Guide (window init refactorings)
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.1.7 for Equalizer 0.4+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.0.1   & Nov 06, 2007 & Fixed Typo\\
    1.1     & Nov 12, 2007 & Adapted distributed objects\\
    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
    1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\subsubsection{Carbon/AGL Thread Safety}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

In direct send compositing, each rendering channel is also responsible
for the sort-last composition\footnote{depth-based for polygonal data or
  sorted-blend for volume data} of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
The frame synchronization through the frame data allows the input frame
to wait for the pixel data to become ready, which is signalled by the
output frame after readback.

@@

Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image. Later versions of Equalizer will
implement asynchronous pixel transfers.

@@

The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
texture-based volume rendering, where the 3D texture is intersected by
view-aligned slices. The slices are rendered back-to-front and blended
together to produce the final image, as shown in
\fig{fSlices}\footnote{Volume Data Set courtesy of: SFB-382 of the German
  Research Council (DFG)}.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
      Volume Rendering}}
\end{wrapfigure}
The full power of scalable volume rendering is however in DB (sort-last)
compounds, where the full volume is divided into separate bricks. Each
of the bricks is rendered like a separate volume. For recomposition, the
\textsf{RGBA} frame buffer data resulting from these render passes then
has to be assembled correctly.

@@

For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
Finding the correct assembly order for perspective frustra is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frustra. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.1.8 for Equalizer 0.4+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.0.1   & Nov 06, 2007 & Fixed Typo\\
    1.1     & Nov 12, 2007 & Adapted distributed objects\\
    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
    1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
    1.1.8   & Jan 15, 2007 & Detailed window initialization based on
                             latest refactoring\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\subsubsection{\label{sCarbonThread}Carbon/AGL Thread Safety}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2.5ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}\vspace{-2ex}}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

\subsection{Advanced Window Initialization}

This section explains window initialization in detail. It discussed in
detail the handling on the different window systems. The entry point for
the window initialization is the task method
\textsf{Window::configInit}. This task method first calls a
window-system-specific task method, and then
\textsf{Window::configInitGL} to do generic OpenGL state setup, as shown
in \fig{fWindowInit}.

Since window initialization is notoriously error-prone and hard to
debug, the default Equalizer functions propagate the reason for errors
from the render clients back to the application. The \textsf{Pipe} and
\textsf{Window} classes have a \textsf{setErrorMessage} method, which is
used to set an error string. This string is passed to the
\textsf{Config} instance on the application node, where it can be
queried using \textsf{getErrorMessage}.

The window-system-specific initialization methods use overrideable
methods for all sub-tasks. This allows partial customization, without
the need of rewriting tedious window initialization code, e.g., the
OpenGL pixel format selection.

\subsubsection{AGL Window Initialization}

AGL initialization happens in three steps: choosing a pixel format,
creating the context and creating a drawable.

Most AGL and Carbon calls are not thread-safe. The Equalizer methods
calling these functions use \textsf{Global::enterCarbon} and
\textsf{Global::leaveCarbon} to protect the API calls. Please refer to
\sref{sCarbonThread} for more details.

The pixel format is chosen based on the window's attributes. Some
attributes set to auto, e.g., stereo, cause the method first to request
the feature and then to back off and retry if it is not available. The
returned pixel format has to be destroyed using
\textsf{Window::destroyAGLPixelFormat}. When no matching pixel format is
found, \textsf{chooseAGLPixelFormat} returns \textsf{0} and the AGL
window initialization returns with a failure.

The context creation also uses the global Carbon lock. Furthermore, it
sets up the swap buffer synchronization with the vertical retrace, if
enabled by the corresponding window attribute hint. Again the window
initialization fails if the context could not be created.

The drawable creation method \textsf{configInitAGLDrawable} calls either
\textsf{configInitAGLFullscreen} or \textsf{configInitAGLWindow}.

The top-level AGL window initialization code therefore looks as follows:

{\footnotesize\begin{lstlisting}
bool Window::configInitAGL()
{
    AGLPixelFormat pixelFormat = chooseAGLPixelFormat();
    if( !pixelFormat )
        return false;

    AGLContext context = createAGLContext( pixelFormat );
    destroyAGLPixelFormat ( pixelFormat );
    setAGLContext( context );

    if( !context )
        return false;

    const bool success = configInitAGLDrawable();
    if( success && !_xDrawable )
    {
        setErrorMessage( "configInitAGLDrawable did set no AGL drawable" );
        return false;
    }

    return success;
}
\end{lstlisting}}

\subsubsection{GLX Window Initialization}

GLX initialization is very similar to AGL initialization. Again the
steps are: choose visual (pixel format), create OpenGL context and then
create drawable. The only difference is that the data returned by
\textsf{chooseXVisualInfo} has to be freed using \textsf{XFree}:

{\footnotesize\begin{lstlisting}
bool Window::configInitGLX()
{
    XVisualInfo* visualInfo = chooseXVisualInfo();
    if( !visualInfo )
        return false;

    GLXContext context = createGLXContext( visualInfo );
    setGLXContext( context );

    if( !context )
        return false;

    const bool success = configInitGLXDrawable( visualInfo );
    XFree( visualInfo );

    if( success && !_xDrawable )
    {
        setErrorMessage( "configInitGLXDrawable did set no X11 drawable" );
        return false;
    }

    return success;    
}
\end{lstlisting}}

\subsubsection{WGL Window Initialization}

The WGL initialization requires another order of operations compared to
AGL or GLX. The following functions are used to initialize a WGL window:

\begin{enumerate}
\item\textsf{getWGLDeviceContext} is used to get a device context which
  might be needed for window creation. Ordinary windows do not need a
  device context during creation, but PBuffers do need one. If a context
  is returned, a function pointer to delete the returned device context
  is set as well.
\item\textsf{configInitWGLDrawable} creates the drawable. The device
  context passed to \textsf{configInitWGLDrawable} is not used when
  creating a window. On success, it sets the window handle.
\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
  window attributes. If no device context is given, it uses the device
  context from this window's window handle. The chosen pixel format is
  set on the window's device context as well as on the passed device
  context. The window needs a window handle set before this method is
  called.
\item\textsf{createWGLContext} creates an OpenGL rendering context using
  the given device context. The device context can not be 0. It does not
  set the window's OpenGL context.
\end{enumerate}

The full \textsf{configInitWGL} task method, including error handling
and cleanup, looks as follows:

{\footnotesize\begin{lstlisting}
bool Window::configInitWGL()
{
    PFNEQDELETEDCPROC deleteDCProc = 0;
    HDC dc = getWGLDeviceContext( deleteDCProc );
    EQASSERT( !dc || deleteDCProc );

    if( !configInitWGLDrawable( dc ))
    {
        if( dc )
            deleteDCProc( dc );
        return false;
    }

    if( !_wglWindowHandle )
    {
        if( dc )
            deleteDCProc( dc );
        setErrorMessage( "configInitWGLDrawable did set no WGL drawable" );
        return false;
    }

    int pixelFormat = chooseWGLPixelFormat( dc );
    if( pixelFormat == 0 )
    {
        if( dc )
            deleteDCProc( dc );
        return false;
    }

    HDC windowDC = GetDC( _wglWindowHandle );
    HGLRC context = createWGLContext( dc ? dc : windowDC );
    ReleaseDC( _wglWindowHandle, windowDC );

    setWGLContext( context );

    if( !context )
    {
        if( dc )
            deleteDCProc( dc );
        return false;
    }

    if( dc )
        deleteDCProc( dc );
    return true;
}
\end{lstlisting}}

@@

The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
In direct send compositing, each rendering channel is also responsible
for the sort-last composition\footnote{depth-based for polygonal data or
  sorted-blend for volume data} of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

The frame synchronization through the frame data allows the input frame
to wait for the pixel data to become ready, which is signalled by the
output frame after readback.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image. Later versions of Equalizer will
implement asynchronous pixel transfers.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
      Volume Rendering}}
\end{wrapfigure}
The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
texture-based volume rendering, where the 3D texture is intersected by
view-aligned slices. The slices are rendered back-to-front and blended
together to produce the final image, as shown in
\fig{fSlices}\footnote{Volume Data Set courtesy of: SFB-382 of the German
  Research Council (DFG)}.

@@

The full power of scalable volume rendering is however in DB (sort-last)
compounds, where the full volume is divided into separate bricks. Each
of the bricks is rendered like a separate volume. For recomposition, the
\textsf{RGBA} frame buffer data resulting from these render passes then
has to be assembled correctly.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

Finding the correct assembly order for perspective frustra is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frustra. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.
